Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-11 Manuscript/manuscript.Rmd
changed def of IR

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo] [??? REFS? Carefully select from my paper notes; !!!added refs]. Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the zinc finger protein, CTCF, has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]-->

[!!! This paragraph defines problems]

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and regions that include TAD boundaries. However, <!--this is a multi-faceted probelem. [??? "WATER"!]-->many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. [??? You confused me a lot with this] Such heavily imbalanced datasets <!--are known to [??? WATER!] -->adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

[!!! The following paragraph should review previous work and highlight pitfalls, creating motivation for yor study]

@@

Class imbalance was characterized by each dataset's imalance ratio (IR), defined as the ratio of minority to majority sets given by the class of genomic bins:

$$IR = \dfrac{C_{minority}}{C_{majority}}$$

where $C_{minority}$ denotes the set of genomic bins that overlapped with flanked TAD boundaries and $C_{majority}$ denotes the set of genomic bins that did not overlap with a flanked TAD boundary. In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling). They are described in detail below.

@@

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently low with very little variability between chromosomes, regardless of cell line.

@@

| Resolution (Bin Width) | Average Number of Genomic Bins (SD) | Average Number of TAD Boundaries (SD) | Average Imbalance Ratio (SD) |
|------------------------|-------------------------------------|---------------------------------------|------------------------------|
| 10 kb                  | 12214.1 (6068.5)                    | 1254.6 (598.8)                        | 0.12 (0.02)                  |  
| 25 kb                  | 4892.7 (2431.2)                     | 714.0 (358.4)                         | 0.17 (0.01)                  |
| 50 kb	                 | 2450.9 (1217.6)                     | 364.4 (184.5)                         | 0.17 (0.01)                  |
| 100 kb                 | 1227.5 (609.8)                      | 165.3 (90.9)                          | 0.15 (0.02)                  |

>>>>

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo] [??? REFS? Carefully select from my paper notes; !!!added refs]. Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the CCCTC-binding factor  (CTCF), has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]-->

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008]. Such heavily imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

Class imbalance was characterized by each dataset's imalance ratio (IR), defined as the ratio of majority to minority sets given by the class of genomic bins:

$$IR = \dfrac{C_{majority}}{C_{minority}}$$

where $C_{majority}$ denotes the set of genomic bins that did not overlap with a flanked TAD boundary and $C_{majority}$ denotes the complementary set of genomic bins that did overlap with a flanked TAD boundary. In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling). They are described in detail below.

@@

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently high with very little variability between chromosomes, regardless of cell line.

@@

| Resolution (Bin Width) | Average Number of Genomic Bins (SD) | Average Number of TAD Boundaries (SD) | Average Imbalance Ratio (SD) |
|------------------------|-------------------------------------|---------------------------------------|------------------------------|
| 10 kb                  | 12214.1 (6068.5)                    | 1254.6 (598.8)                        | 8.7 (1.2)                    |  
| 25 kb                  | 4892.7 (2431.2)                     | 714.0 (358.4)                         | 5.9 (0.5)                    |
| 50 kb	                 | 2450.9 (1217.6)                     | 364.4 (184.5)                         | 5.8 (0.5)                    |
| 100 kb                 | 1227.5 (609.8)                      | 165.3 (90.9)                          | 6.8 (1.1)                    |
